Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-29 numerical-chemistry.tex

While it is often safe to assume that species (both chemical and
ionization) within a gas can be treated as being in equilibrium, in
some regimes that are found in astrophysics this assumption leads to
considerable error.  For example, the cooling and collapse of
primordial gas in Population III star formation is dominated by
molecular hydrogen, which in the absence of dust forms via an
inefficient pair of collisional processes that depend heavily on the
local population of free electrons.  As a result, when modeling
primordial star formation, it is critical to follow the
non-equilibrium evolution of the chemical species of hydrogen,
including molecular hydrogen and deuterium.

The primordial non-equilibrium chemistry routines used in \enzo\ were
first described by Abel et al. and Anninos et
al. \citep{abel97,anninos97}, but have been extended since then with
updated reaction rates and the inclusion of deuterium species
\citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12
total species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp,
and HD.  (A separate, but related, module follows the radiative
heating and cooling of the gas from atomic and molecular line
excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as
well as several models for a metagalactic UV background that heat the
gas via photoionization and photodissociation; see
Section~\ref{sec.num.cooling} for more details).  Depending on user
parameters, one can follow only atomic species (H, \Hp, He, \Hep,
\Hepp, and e$^-$), include those species relevant for molecular
hydrogen formation (\HH, \HHp, and \Hm), and can further include
deuterium and its species (D, \Dp, and HD).  A total of 9 kinetic
equations are solved, including 29 kinetic and radiative processes,
for the 12 species mentioned above.  See Table~\ref{table.collisional}
for the collisional processes and Table~\ref{table.radiative} for the
radiative processes solved.

The chemical reaction equation network is technically challenging to
solve due to the huge range of reaction timescales involved -- the
characteristic creation and destruction timescales of the various
species and reactions can differ by many orders of magnitude.  As a
result, the set of rate equations is extremely stiff, and an explicit
scheme for integration of the rate equations can be exceptionally
costly if small enough timesteps are taken to keep the network stable.
This makes an implicit scheme strongly preferable for such a set of
equations.  However, an implicit scheme typically requires an iterative
procedure to converge, and for large networks (such as this one) an
implicit method can be very time-consuming, making it undesirable for
a large, three-dimensional simulation.

\enzo\ solves the rate equations using a method based on a first-order-accurate, implicit
Euler formulation in order to provide a stable and accurate
solution. This technique is optimized by taking the chemical
intermediaries \Hm and \HHp, which have large rate coefficients and
low concentrations, and grouping them into a separate category of
equations.  Due to their fast reactions, these species are very
sensitive to small changes in the more abundant species.  However, due
to their low overall concentrations, they do not significantly
influence the abundance of species with higher concentrations.
Therefore, reactions involving these two species can be decoupled from
the rest of the network and treated independently.  In fact, \Hm~and
\HHp~are treated as being in equilibrium at all times, independent of
the other species and the hydrodynamic variables.  This allows a large
speedup in solution as the implicit Euler scheme is then applied only to the
slower 7- or 10-species network (depending on whether deuterium is
included or not) on timescales closer to those required by the
hydrodynamics of the simulation.  Even so, the accuracy and stability
of the scheme is maintained by subcycling the rate solver over a
single hydrodynamic timestep.  These subcycle timesteps are determined
so that the estimated fractional change in the electron concentration
is limited to no more than $10\%$ per timestep.  Given this
subcycling, the method is not truly implicit over an entire
hydrodynamical timestep  -- however,
it uses the formalism and is implicit over subcycle timesteps.

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling eventually breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$ in gas of primordial composition.  Beyond this point,
modifications to the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.  Even
with these modifications, a completely correct description of the
cooling of primordial gas requires some form of radiation transport,
which will greatly increase the cost of the simulations.

>>>>

While it is often safe to assume that species (both chemical and ionization)
within a gas can be treated as being in equilibrium, in some regimes that are
found in astrophysics this assumption leads to considerable error.  For
example, the cooling and collapse of primordial gas in Population III star
formation is dominated by molecular hydrogen, which in the absence of dust
forms via an inefficient pair of collisional processes that depend heavily on
the local, highly non-equilibrium population of free electrons.  As a result,
when modeling primordial star formation, it is critical to follow the
non-equilibrium evolution of the chemical species of hydrogen, including
molecular hydrogen and deuterium.

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by Abel et al. and Anninos et al. \citep{abel97,anninos97}, but have
been extended since then with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  Enzo
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation; see Section~\ref{sec.num.cooling} for
more details.  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., de-coupled and operator split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that faster than those of the gas dynamics.

Input parameters to \enzo\ govern the chemical species that are updated during
the course of the simulation.  This can include only the atomic species (H,
\Hp, He, \Hep, \Hepp, and e$^-$), those species relevant for molecular hydrogen
formation (\HH, \HHp, and \Hm), and can further include deuterium and its
species (D, \Dp, and HD).  A total of 9 kinetic equations are solved, including
29 kinetic and radiative processes, for the 12 species mentioned above.  See
Table~\ref{table.collisional} for the collisional processes and
Table~\ref{table.radiative} for the radiative processes solved.

The chemical reaction equation network is technically challenging to solve due
to the huge range of reaction timescales involved.  The characteristic times
for creation and destruction of the various species and reactions can differ by
many orders of magnitude and are often very sensitive to the chemical and
thermal state of the gas.  This makes a fully-implicit scheme, with convergence
criteria and error tolerance, strongly preferable for such a set of equations.
However, most implicit schemes require an iterative procedure to converge, and
for large networks (such as this one) an iterative, fully-implicit method can
be very time-consuming and computationally costly for a relatively small
increase in accuracy.  At this present time, this makes fully-implicit methods
somewhat undesirable for a large, three-dimensional simulation.

\enzo\ solves the rate equations using a method based on a semi-implicit
formulation in order to provide a stable, positive definite and first-order
accurate solution.  The update discretization splits chemical changes into
formation components and destruction components and updated with a mixed set of
time states, as described in \citet{anninos97}.  The formation components of
species $S_i$ are computed at the current subcycle time, where the contribution
of species $S_i$ to its own destruction components are computed at the updated
time; all other contributions to the destruction component are computed at the
current time.  This mixed state improves accuracy and ensures species values
are positive definite, and is equivalent to one Jacobi iteration of an implicit
Euler solve.  This technique is optimized by taking the chemical intermediaries
\Hm and \HHp, which have large rate coefficients and low concentrations, and
grouping them into a separate category of equations.  Due to their fast
reactions, these species are very sensitive to small changes in the more
abundant species and are at all times in astrophysical calculations close to
equilibrium values.  Attempting to resolve their formation and destruction
times would necessitate extremely small timesteps, and while these species
provide necessary channels to facilitate the formation of molecular hydrogen at
low densities, they do not otherwise significantly influence the abundance of
species with higher concentrations.  Therefore, reactions governing these two
species can be decoupled from the rest of the network and treated independently
through analytic solutions for equilibrium values.  This allows a large speedup
in solution as the timestepping scheme is applied only to the slower 7- or
10-species network (depending on whether deuterium is included or not), which
will be much closer to the overall hydrodynamic timestep of the simulation.

Even so, the accuracy and stability of the scheme is maintained by subcycling
the rate solver within a single hydrodynamic timestep.  These subcycle
timesteps are determined so that the estimated fractional change in the
electron concentration is limited to no more than $10\%$ per timestep;
additional criteria may be applied based on the expected change from radiative
cooling, chemical heating from the formation of molecular hydrogen.

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling eventually breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$ in gas of primordial composition.  Beyond this point,
modifications to the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.   The formation
of molecular hydrogen as catalyzed by dust was recently added to \enzo\ to
enable studies of low-metallicity gas, as well as the inclusion of appropriate
timestepping criteria to account for the input of ionizing radiation.  Even
with these modifications, a completely correct description of the cooling of
primordial gas at very high densities requires some form of radiation
transport, which will greatly increase the cost of the simulations.
Furthermore, at very high densities, the stiffness of the molecular hydrogen
reaction rates may require better than a first-order accurate solution; as
such, the transition to this regime will likely necessitate a fully-implicit,
iterative solver.

### 2013-05-29 numerical-radiation-fld.tex

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR) grids, which has been tuned for large-scale simulations
involving many ionizing sources.  Detailed explanations of the model and
solution approach may be found in
\cite{NBHBROW2007}, \cite{ReynoldsHayesPaschosNorman2009} and
\cite{NRS2009}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.

>>>>

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR, non-SMR) grids, which has been tuned for large-scale simulations
involving many ionizing sources.  Detailed explanations of the model and
solution approach may be found in
\cite{NBHBROW2007}, \cite{ReynoldsHayesPaschosNorman2009} and
\cite{NRS2009}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.

### 2013-05-29 tests-doublemach.tex

The double mach reflection test is a classic two-dimensional test of
hydrodynamic algorithms, and was originally described in
\citet{1984JCoPh..54..115W} \citep[and more recently
in][]{2008ApJS..178..137S}.  In this problem, shown in
Figure~\ref{fig.doublemach}, a shock is injected at an angle to a
reflecting surface (the -y boundary), and a jet appears along the
reflecting surface.  The ideal solution is self-similar, and the
appearance of this solution is highly sensitive to numerical
diffusion.  If numerical noise is present, a Kelvin-Helmholz
instability develops along this jet and breaks the self-similarity.

In the test problem shown in Figure~\ref{fig.doublemach}, a 2D
simulation with $960 \times 240$ cells was formed with a domain of $x
= [0, 4]$ and $y = [0, 1]$.  We use an ideal gas equation of state of
$\gamma = 1.4$, a pre-shock density of 1.4, and a pre-shock specific
internal energy of $2.5/1.4$ (all in arbitrary units).  A Mach 10
shock is initialized with a shock normal that is 30$^\circ$ from the
x-axis and an initial position on the lower boundary of x$ = 1/6$.
The lower y boundary and right x boundary are reflecting; the left x
boundary is inflowing, and the upper y boundary has a time-dependent
boundary condition that allows the shock to propagate into the domain
as if it extends to infinity.  The simulation starts at t$ = 0$ and
runs until t$ = 0.205$ (arbitrary units), at which point the rightmost
extent of the shock should be at roughly x$ = 3$.  In this run, we use
the direct Eulerian implementation of the piecewise parabolic hydro
method, with the diffusion, flattening, and shock steepening all
turned on.

>>>>

The double mach reflection test is a classic two-dimensional test of
hydrodynamic algorithms originally described in
\citet{1984JCoPh..54..115W} \citep[and more recently
in][]{2008ApJS..178..137S}.  In this problem, shown in
Figure~\ref{fig.doublemach}, a shock is injected at an angle to a
reflecting surface (the -y boundary), and a jet appears along the
reflecting surface.  The ideal solution is self-similar, and the
appearance of this solution is highly sensitive to numerical
diffusion.  If numerical noise is present, a Kelvin-Helmholz
instability develops along this jet and breaks the self-similarity.

In the test problem shown in Figure~\ref{fig.doublemach}, a 2D
simulation with $960 \times 240$ cells was created with a domain of $x
= [0, 4]$ and $y = [0, 1]$.  We use an ideal gas equation of state of
$\gamma = 1.4$, a pre-shock density of 1.4, and a pre-shock specific
internal energy of $2.5/1.4$ (all in arbitrary units).  A Mach 10
shock is initialized with a shock normal that is 30$^\circ$ from the
x-axis and an initial position on the lower boundary of x$ = 1/6$.
The lower y boundary and right x boundary are reflecting; the left x
boundary is inflowing, and the upper y boundary has a time-dependent
boundary condition that allows the shock to propagate into the domain
as if it extends to infinity.  The simulation starts at t$ = 0$ and
runs until t$ = 0.205$ (arbitrary units), at which point the rightmost
extent of the shock should be at roughly x$ = 3$.  In this simulation we use
the direct Eulerian implementation of the piecewise parabolic hydrodynamic
method with the diffusion, flattening, and shock steepening all enabled.

### 2013-05-29 tests-vandv.tex

Several comparisons have been made that focus specifically on
hydrodynamics solvers and fluid behavior.  In particular, the work of
\citet{2007MNRAS.380..963A} and \citet{Tasker2008} perform
head-to-head comparisons between several grid- and particle-based
codes for a variety of fluid-centric test problems (including shocked gas clouds,
self-gravitating, translating clouds, and Sedov-Taylor explosions),
and show that \enzo\ is comparable or superior in behavior to the
other grid-based hydrodynamics codes involved in the comparison, and
provide useful information on the sort of practical challenges that a
user of an
AMR code such as \enzo\ may experience.  More specific comparisons, including one
specifically testing the linear and nonlinear growth of the
Kelvin-Helmholz instability \citep{2012ApJS..201...18M}, as well as a
comparison that more
broadly examines Galilean invariance in grid-based codes
\citep{2010MNRAS.401.2463R}, show that \enzo, and in particular its
implementation of the Piecewise Parabolic Method hydrodynamics solver,
converge to the correct solution as expected, and generally provide
less diffusive solutions than lower-order codes, including those that
use artificial viscosity.  Finally, two code comparison projects have
been undertaken to examine turbulence in hydro codes -- one studied
the behavior of decaying isothermal supersonic turbulence
\citep{2009A&A...508..541K}, and the other looked at supersonic
magnetohydrodynamical turbulence \citep{2011ApJ...737...13K}.  Both
included the \enzo\ code, with the former testing the PPM
hydrodynamics and the latter the constrained transport MHD
implementation of \citet{Collins10}.  In both cases, \enzo\ performed similarly to other
grid-based codes that use Godunov-based fluid solvers, and typically
had better effective resolution than particle-based codes using the
same number of particles as the \enzo\ simulation used grid cells.

Three other comparisons between the \enzo\ code and other simulation
tools have been performed that focus on physics other than gravity and
fluid flow.  The flux-limited diffusion radiation transport scheme was
measured against several test problems by \citet{IlievEtAl2009},
producing results that are similar to both expected results (i.e.,
analytic test problems) and other methods (i.e., codes using other
methods, written by other groups), although there are some minor
differences (although all codes in the study exhibited differences
from the majority of other codes on at least a subset of the tests).
\citet{2011ApJ...726...55T} shows the results of varied reaction rate
coefficients for the formation of molecular hydrogen via the 3-body
process in both the \enzo\ and Gadget-2 codes.  Similar trends were
visible in both the \enzo\ and Gadget simulations, though at a
nominally equivalent resolution (i.e., with particle and cell gas
masses being comparable) \enzo\ simulations typically displayed a
substantially higher level of gas structure.  This is unsurprising due
to \enzo's higher-order hydro solver.  Finally,
\citet{2012ApJ...744...52P} show the results of comparing \enzo\ in
its non-AMR mode to a smoothed-particle hydrodynamics code, SNSPH, in
the context of common-envelope binary stellar interactions.  The
authors show that the codes display reasonable convergence properties
as a function of simulation resolution, and also agree quite well with
each other -- however, the observed mass-loss rates do not agree
particularly well with observations.

>>>>

Several comparisons have been made that focus specifically on
hydrodynamics solvers and fluid behavior.  In particular, the work of
\citet{2007MNRAS.380..963A} and \citet{Tasker2008} perform
direct comparisons between several grid- and particle-based
codes for a variety of fluid-centric test problems (including shocked gas clouds,
self-gravitating, translating clouds, and Sedov-Taylor explosions),
and show that \enzo\ is comparable or superior in behavior to the
other grid-based hydrodynamics codes involved in the comparison, and
provide useful information on the sort of practical challenges that a
user of an
AMR code such as \enzo\ may experience.  More specific comparisons, including one
specifically testing the linear and nonlinear growth of the
Kelvin-Helmholz instability \citep{2012ApJS..201...18M}, as well as a
comparison that more
broadly examines Galilean invariance in grid-based codes
\citep{2010MNRAS.401.2463R}, show that \enzo, and in particular its
implementation of the Piecewise Parabolic Method hydrodynamics solver,
converge to the correct solution as expected, and generally provide
less diffusive solutions than lower-order codes, including those that
use artificial viscosity.  Finally, there have been two code comparison
projects that focus on turbulence simulations.  The first studied the
behavior of decaying isothermal supersonic turbulence
\citep{2009A&A...508..541K}, and the second examined supersonic
magnetohydrodynamical turbulence \citep{2011ApJ...737...13K}.  Both
included the \enzo\ code, with the former testing the PPM
hydrodynamics and the latter the constrained transport MHD
implementation of \citet{Collins10}.  In both cases, \enzo\ performed similarly to other
grid-based codes that use Godunov-based fluid solvers, and typically
had better effective resolution than particle-based codes when using the
same number of particles as the number of grid cells in the \enzo\ simulation.

Three other comparisons between the \enzo\ code and other simulation
tools have been performed that focus on physics other than gravity and
fluid flow.  The flux-limited diffusion radiation transport scheme was
measured against several test problems by \citet{IlievEtAl2009},
which involved tests with and without analytical solutions.  \enzo\ produced
results similar to both the analytical solutions and results obtained by 
other codes.  We note, however, that there were minor differences throughout
the comparison between various codes, and the majority of the codes differred
in at least a subset of the tests.
\citet{2011ApJ...726...55T} shows the results of varied reaction rate
coefficients for the formation of molecular hydrogen via the 3-body
process in both the \enzo\ and Gadget-2 codes, observing similar trends between
the two codes. However, at nominally equivalend resolution
where the particle and cell gas masses are comparable, 
\enzo\ simulations typically displayed a
substantially higher level of gas structure.  This is unsurprising due
to \enzo's higher-order hydro solver.  Finally,
\citet{2012ApJ...744...52P} show the results of comparing \enzo\ in
its non-AMR mode to a smoothed-particle hydrodynamics code, SNSPH, in
the context of common-envelope binary stellar interactions.  The
authors show that the codes display reasonable convergence properties
as a function of simulation resolution, and also agree quite well with
each other. However, the observed mass-loss rates do not agree
particularly well with observations.

### 2013-05-29 tests-wavepool.tex

In this one-dimensional test, we pass a short wavelength, linear wave through a static refined region.  The full domain is from 0 to 1 and the refined region is set from 0.25 to 0.75 at all times.  We modify the left boundary consistent with a single linear sound wave with density given by $\rho(x,t) = \rho_0 (1 + A \sin(kx - \omega t))$, where the amplitude is $A = 0.01$ and the wavelength $\lambda = k/2\pi = 0.1$, and similar expressions exist for the pressure and velocity.  The initial, unperturbed density and pressure are set to unity and we adopt $\gamma = 1.4$.  The (unrefined) root grid is covered with 100 cells, resulting in a wavelength for the linear wave of only 10 cells.  This is, therefore, a challenging problem for hydro methods.  We are particularly interested in any reflection or artifacts introduced by the wave entering and exiting the refined regions.

Figure~\ref{fig.wavepool} shows the evolution of the wave at three different times ($t = 0.2, 0.3$ and 0.8) for three of our solvers (PPM, Zeus, and MUSCL).  The leftmost column show the wave just before it enters the refined region so that we can gauge how the solver is operating in the absence of AMR, the center column shows the wave after it has fully entered the refined region, and the rightmost column shows the wave well after it has exited the refined region.

Beginning with the PPM, we see that even before it has entered the refine region, the wave is slightly damped, which is not unexpected for such a short wavelength mode, even for PPM.  The remaining panels demonstrate that the wave cleanly enters and exits the refined region.  No significant reflection is seen, either on entry or exit, and the amount of damping is mild.

For the Zeus solver, we see that even before it has entered the refined region, there are small oscillations excited behind the wave; although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test, we use only our standard, low amount of quadratic artificial viscosity -- these waves can be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise -- the end result is similar to the case without any refined region.

Finally, the MUSCL solver shows also shows a very clean entry and exit from the refined region without any oscillations, although (for the PLM reconstruction used here) the wave is spread more than with the other methods.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/WavePool.eps}
\caption{This plot shows, for each column, three snap shots (at $t=0.2, 0.3$ and 0.8) of a linear wave as it propagates through the grid.  The static refined region extends from $x = 0.25$ to 0.75 and is shown in grey in each panel.  The individual cells are also color-coded by level: blue indicates the root grid, and red is for the refined region.  Each row shows the result for a different solver (top is PPM, middle is Zeus, and the bottom is MUSCL).   The solid line shows the analytic solution for a linear, undamped wave.  Note that we focus each panel on a small region of the entire domain to better show the wave itself.}
\label{fig.wavepool}
\end{center}
\end{figure}

>>>>

In this one-dimensional test we pass a short wavelength linear wave through a static refined region.  The full domain is from 0 to 1 and the refined region is set from 0.25 to 0.75 at all times.  We modify the left boundary consistent with a single linear sound wave with density given by $\rho(x,t) = \rho_0 (1 + A \sin(kx - \omega t))$, where the amplitude is $A = 0.01$ and the wavelength $\lambda = k/2\pi = 0.1$, and similar expressions exist for the pressure and velocity.  The initial, unperturbed density and pressure are set to unity and we adopt $\gamma = 1.4$.  The (unrefined) root grid is covered with 100 cells, resulting in a wavelength for the linear wave of only 10 cells.  This is, therefore, a challenging problem for hydro methods.  We are particularly interested in any reflection or artifacts introduced by the wave entering and exiting the refined regions.

Figure~\ref{fig.wavepool} shows the evolution of the wave at three different times ($t = 0.2, 0.3$ and 0.8) for three of our solvers (PPM, Zeus, and MUSCL).  In all panels, the shaded regions denote the statically refined region. The leftmost column show the wave just before it enters the refined region so that we can gauge how the solver is operating in the absence of AMR.  The center column shows the wave after it has fully entered the refined region.  The rightmost column shows the wave well after it has exited the refined region.

With the PPM solver, we notice that even before the wave reaches the refined region it has been slightly damped. This is not unexpected for such a short wavelength mode, even for high-order solvers such as PPM.  The remaining panels demonstrate that the wave cleanly enters and exits the refined region.  No significant reflection is seen on either entry or exit, and the amount of damping is mild.

For the Zeus solver, we see that even before it has entered the refined region there are small oscillations excited behind the wave; although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test we use only our standard, low amount of quadratic artificial viscosity.  These waves could be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise. We note that the end result is similar to the case without any refined region.

Finally, the MUSCL solver shows also shows a very clean entry and exit from the refined region without any oscillations. However,
because we use a piecewise linear reconstruction, the wave is spread more than with the other methods.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/WavePool.eps}
\caption{This plot shows, for each column, three snapshots at $t=0.2, 0.3$ and 0.8 of a linear wave as it propagates through the domain.  A static refined region extends from $x = 0.25$ to 0.75 and is shown in grey in each panel.  The individual cells are also color-coded by level: blue indicates the root grid, and red is for the refined region.  Each row shows the result for a different hydrodynamic solver (top is PPM, middle is Zeus, and the bottom is MUSCL).   The solid line shows the analytic solution for a linear, undamped wave.  Note that we focus each panel on a small region of the entire domain to better show the wave itself.}
\label{fig.wavepool}
\end{center}
\end{figure}
